---
title: "DATA 612 Project 3"
author: "Brian Liles"
date: "June 25, 2019"
output: html_document
---
#Matrix Factorization Methods

Singular Decomposition in R
$$A = U \Sigma V^T$$

According to Wikipedia, the **singular value decomposition** is a factorization of a real or complex matrix. In order to conduct this tasks the data must be complete.
```{r}
# create a matrix entitled a
a <- matrix(c(4,8,3,8,7,
              3,1,1,0,2,
              1,1,1,0,6,
              0,0,1,7,0,
              3,7,6,8,0,
              2,4,3,5,3), nrow = 6, byrow = TRUE)

# create colnames
colnames(a) <- c("The Wire","Breaking Bad","The Sopranos","Game of Thrones","The Shield")

# create rownames
rownames(a) <- c("User1","User2","User3","User4","User5","User6")
a
```

```{r}
# obtain the rank of the matrix 
qr(a)$rank
```

```{r}
# conduct the svd of the matrix a
svd_a <- svd(a)
```

Column **U** is the eigenvector of $AA^T$
```{r}
# view column U of matrix a
svd_a$u
```

Column **V** is the eigenvector of $A^TA$
```{r}
# view column V of matrix a
svd_a$v
```

The diagonal matrix $\Sigma$ are the square roots of the non-zero eigenvalues of $AA^T$ and $A^TA$
```{r}
# view the diagonal matrix
(d <- diag(svd_a$d))
```

#Test and Training Data Sets
```{r}
set.seed(50)
sample <- sample.int(n = nrow(a), size = floor(.75*nrow(a)), replace = F)
aTrain <- a[sample,]
aTest <- a[-sample,]
```

#Conduct SVD for the training and test data sets
```{r}
# conduct the svd of the matrix a
(svd_aTrain <- svd(aTrain))
```

```{r}
# conduct the svd of the matrix a
(svd_aTest <- svd(aTest))
```